Abstraction and Invariance for Algebraically Indexed Types

### 2012-10-26 paper/section-instantiations.tex
Various little fixes to Sections 1,2,4 and 5
Section 1:
- Make the diagram in Figure 1 a little tighter, and add a reference
  in the text (in the fourth bullet point in the contributions).

Section 2:
- Introduce to the \floor and semantic brackets explicitly in Section
  2.

- Change the brackets around the intended semantics of affComb

- Insert a line break after affComb in Figure 3 and remove the
  semantics of affComb.

Section 4:
- Fix a small typo in the type isomorphism for scalings ("A[1/s]"
  instead of just "[1/s]")

Section 5:
- Two small wording improvements
Put commas after all

Unfolding definitions, we deduce that
\[
\tmSem{I(J(y))}(\opsEnv,y) = \tmSem{\transup A t v}(\opsEnv,\vec 0)(y)
\]
for any $y \in \tmSem{A[0/t]}$. It's easy to prove by induction on $A$
that $\tmSem{\transup A t v}(\opsEnv,\vec 0)(y) = y$, and so
$\ctxteq\DeltaWithB{y\mathord:Y}{J(I(y))}y Y$ follows.

@@

\exref{ex:type-iso} and \exref{ex:simpletrans} are special cases of
this isomorphism. Another instance is the type of vector addition:
\begin{align*}
\forall B\mathord:\SynGL{2}, t_1, t_2\mathord:\SynTransl{2}.\tyPrim{vec}{B,t_1}\to\tyPrim{vec}{B,t_2}\to\tyPrim{vec}{B,t_1\mathord+t_2}\\
\cong\quad\forall B\mathord:\SynGL{2}, t_1\mathord:\SynTransl{2}.\tyPrim{vec}{B,t_1}\to\tyPrim{vec}{B,t_1}
\quad (\text{by~\lemref{lem:transiso}})\\
\cong\quad\forall B\mathord:\SynGL{2}.\tyPrim{vec}{B,0} \quad (\text{by~\lemref{lem:transiso}})\\
\cong\quad\tyUnit\quad(\text{by \autoref{thm:abstraction}})
\end{align*}
In other words, the vector addition operation is the only inhabitant of its type!
\paragraph{Scalings}
For the group $\GL{1}$ of scalings we can treat
a real-valued argument to a function as a unit-of-measure with
which to scale the result. 
%The small fly in the ointment is that the
Although the argument might be zero, and this cannot be used for scaling,
%Nevertheless, 
we still can obtain the following slightly more complicated
isomorphism:
\begin{eqnarray}
\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to A &
\cong & [1/s]\times \forall s\mathord:\SynGL{1}.A\label{eqn:scaleiso}
\end{eqnarray}
The type of $\mathrm{areaCircle}$ from the introduction is one instance:
\begin{align*}
\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to\tyPrim{real}{s^2}\\ 
\cong\quad\tyPrim{real}{1}\times(\forall
s\mathord:\SynGL{1}.\tyPrim{real}{s^2}) \quad(\text{by \autoref{eqn:scaleiso}})\\ 
\cong\quad\tyPrim{real}{1}\times \tyUnit \quad (\text{by \autoref{thm:abstraction}})\\
\cong\quad\tyPrim{real}{1}\quad (\text{trivially})
\end{align*}
%where %the isomorphism 
%The abstraction theorem is used to prove
%$\forall s\mathord:\SynGL{1}.\tyPrim{real}{s^2}\cong \tyUnit$.
% is proved using
%the abstraction theorem.
%\begin{enumerate}
%\item Two cases: when we have a total ``division'' operation, and when
%  we don't.
%\item Type isomorphisms, based on Smith Normal Form
% \item Non-definability results, based on the non-compositional
%   relational environments
% \item Alteration of the non-definability results in the light of
%   addition of a square root operation
% \item Mention the variation we can obtain by changing some types,
%   e.g. the type of $\mathrm{abs}$. Also, the complex numbers example.
%\end{enumerate}

@@

\newcommand{\Grp}{\mathit{Gr}}
To prove non-definability results, the model and relational interpretation
used in the previous two sections are not sufficient.
This is true even for simple scalars with invariance under scaling.
Consider the type 
$\forall s \mathord: \SynGL{1}.\tyPrim{real}{s^2} \to \tyPrim{real}{s}$. 
There are many functions~$f: \mathbb R \to \mathbb R$
that satisfy its relational interpretation, i.e. for which $f(k^2\cdot x) = k\cdot f(x)$ for 
any $k\neq 0$.
Consider 
\[
f(x) = \begin{cases}\sqrt x & \text{if }x > 0\\0&\text{otherwise}\end{cases}
\]
for instance. Yet this type contains only the constant zero function, a fact that
we can prove using a surprisingly simple model!

@@

You can think of $z$ as tracking whether or not exponents on scale
parameters are divisible by 2: all the primitive operations in
\autoref{fig:real-ops} produce results with even exponents if their
inputs have even exponents. But the type under consideration here does
not.

@@

Here $q$ is tracking the exponents on scale parameters, with
$\semPrimType{\tyPrimNm{real}}(q)$ relating only zero to itself
unless the denominator of $q$ is a power of two (a so-called
\emph{dyadic} number).  
Applying the abstraction theorem and expanding, we have that for
any
$q\in\mathbb Q$ and any
$(x,y)\in\semPrimType{\tyPrimNm{real}}(3q)$, it is the case that
$(f(x),f(y))\in\semPrimType{\tyPrimNm{real}}(q)$. Set $q = \frac 1 3$.  
Then we have that if $x=y$ then $f(x)=f(y)=0$.
\qed
\end{example}

>>>>

Unfolding definitions, we deduce that
\[
\tmSem{I(J(y))}(\opsEnv,y) = \tmSem{\transup A t v}(\opsEnv,\vec 0)(y)
\]
for any $y \in \tmSem{A[0/t]}$. It's easy to prove by induction on $A$
that $\tmSem{\transup A t v}(\opsEnv,\vec 0)(y) = y$, and so
$\ctxteq\DeltaWithB{y\mathord:Y}{I(J(y))}y Y$ follows.

@@

\exref{ex:type-iso} and \exref{ex:simpletrans} are special cases of
this isomorphism. Another instance is the type of vector addition:
\begin{align*}
\forall B\mathord:\SynGL{2}, t_1, t_2\mathord:\SynTransl{2}.\tyPrim{vec}{B,t_1}\to\tyPrim{vec}{B,t_2}\to\tyPrim{vec}{B,t_1\mathord+t_2}\\
\cong\quad\forall B\mathord:\SynGL{2}, t_1\mathord:\SynTransl{2}.\tyPrim{vec}{B,t_1}\to\tyPrim{vec}{B,t_1}
\quad (\text{by~\lemref{lem:transiso}})\\
\cong\quad\forall B\mathord:\SynGL{2}.\tyPrim{vec}{B,0} \quad (\text{by~\lemref{lem:transiso}})\\
\cong\quad\tyUnit\quad(\text{by \autoref{thm:abstraction}})
\end{align*}
In other words, the vector addition operation is the only inhabitant of its type!
\paragraph{Scalings}
For the group $\GL{1}$ of scalings we can treat
a real-valued argument to a function as a unit-of-measure with
which to scale the result. 
%The small fly in the ointment is that the
Although the argument might be zero, and this cannot be used for scaling,
%Nevertheless, 
we still can obtain the following slightly more complicated
isomorphism:
\begin{eqnarray}
\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to A &
\cong & A[1/s]\times \forall s\mathord:\SynGL{1}.A\label{eqn:scaleiso}
\end{eqnarray}
The type of $\mathrm{areaCircle}$ from the introduction is one instance:
\begin{align*}
\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to\tyPrim{real}{s^2}\\ 
\cong\quad\tyPrim{real}{1}\times(\forall
s\mathord:\SynGL{1}.\tyPrim{real}{s^2}) \quad(\text{by \autoref{eqn:scaleiso}})\\ 
\cong\quad\tyPrim{real}{1}\times \tyUnit \quad (\text{by \autoref{thm:abstraction}})\\
\cong\quad\tyPrim{real}{1}\quad (\text{trivially})
\end{align*}
%where %the isomorphism 
%The abstraction theorem is used to prove
%$\forall s\mathord:\SynGL{1}.\tyPrim{real}{s^2}\cong \tyUnit$.
% is proved using
%the abstraction theorem.
%\begin{enumerate}
%\item Two cases: when we have a total ``division'' operation, and when
%  we don't.
%\item Type isomorphisms, based on Smith Normal Form
% \item Non-definability results, based on the non-compositional
%   relational environments
% \item Alteration of the non-definability results in the light of
%   addition of a square root operation
% \item Mention the variation we can obtain by changing some types,
%   e.g. the type of $\mathrm{abs}$. Also, the complex numbers example.
%\end{enumerate}

@@

\newcommand{\Grp}{\mathit{Gr}}
To prove non-definability results, the model and relational interpretation
used in the previous two sections are not sufficient.
This is true even for simple scalars with invariance under scaling.
Consider the type 
$\forall s \mathord: \SynGL{1}.\tyPrim{real}{s^2} \to \tyPrim{real}{s}$. 
There are many functions~$f: \mathbb R \to \mathbb R$
that satisfy its relational interpretation, i.e., for which $f(k^2\cdot x) = k\cdot f(x)$ for 
any $k\neq 0$.
Consider 
\[
f(x) = \begin{cases}\sqrt x & \text{if }x > 0\\0&\text{otherwise}\end{cases}
\]
for instance. Yet this type contains only the constant zero function, a fact that
we can prove using a surprisingly simple model!

@@

We think of $z$ as tracking whether or not exponents on scale
parameters are divisible by 2: all the primitive operations in
\autoref{fig:real-ops} produce results with even exponents if their
inputs have even exponents. But the type under consideration here does
not.

@@

Here, $q$ is tracking the exponents on scale parameters, with
$\semPrimType{\tyPrimNm{real}}(q)$ relating only zero to itself
unless the denominator of $q$ is a power of two (a so-called
\emph{dyadic} number).  
Applying the abstraction theorem and expanding, we have that for
any
$q\in\mathbb Q$ and any
$(x,y)\in\semPrimType{\tyPrimNm{real}}(3q)$, it is the case that
$(f(x),f(y))\in\semPrimType{\tyPrimNm{real}}(q)$. Set $q = \frac 1 3$.  
Then we have that if $x=y$ then $f(x)=f(y)=0$.
\qed
\end{example}
